Model-based Clustering for Neural Populations

### 2021-09-08 documents/cluster_neural_population.tex
new documents

For completeness, Gibbs samplers for all three versions (no constraint, block diagonal and diagonal \(\mathbf{Q}\)) are shown below.

@@

The conditional priors for these parameters:
\begin{enumerate}
	\def\labelenumi{(\arabic{enumi})}
	\item
	Latent vectors \(\left\{ \mathbf{x}_{t} \right\}_{t=1}^T\): the conditional prior is
	defined by
	\[\mathbf{x}_{1} \sim N(\mathbf{x}_{0},\mathbf{Q}_{0})\]
	\[\mathbf{x}_{t + 1}|\mathbf{x}_{t} \sim N(\mathbf{A}\mathbf{x}_{t} + \mathbf{b},\mathbf{Q})\]
	\item
	Initials \(\mathbf{x}_{0}\): assume there are \(J\) clusters,
	\[\mathbf{x}_{0} \sim N(\bm{\mu}_{\mathbf{x}_{00}},\ \mathbf{\Sigma}_{\mathbf{x}_{00}})\]
	, where \(\bm{\mu}_{\mathbf{x}_{00}} = \mathbf{0}_{\text{Jp}}\) and \(\mathbf{\Sigma}_{\mathbf{x}_{00}} = \mathbf{I}_{\text{Jp}}\)
	\item
	Linear mapping (loading) for latent vectors

	\textcolor{red}{\textbf{I just ignore that it is just multivariate GLM problem. Do it later...}}\\

	\item
	Mean and covariance for loading in each cluster

	\textcolor{red}{\textbf{MV-GLM...}}\\

	\item
	Linear dynamics for latent vectors \(\mathbf{A}\) and \(\mathbf{b}\): if the number of cluster is \(J\)

	\textcolor{red}{\textbf{MV-GLM...}}\\

	\item
	Process noise \(\mathbf{Q}\): if the number of cluster is \(J\)

	\textcolor{red}{\textbf{MV-GLM...}}\\

@@

\subsubsection{Update \(\left\{ d_{i} \right\}_{i = 1}^{N}\) and
	\(\left\{ \mathbf{c}_{i} \right\}_{i = 1}^{N}\)}
To update efficiently, use Laplace approximation again. Denote

\textcolor{red}{\textbf{MV-GLM...}}\\

@@

\textcolor{red}{\textbf{MV-GLM...}}\\

\subsubsection{Update \(\mathbf{A}\) and \(\mathbf{b}\)} \label{dynamics update}

\textcolor{red}{\textbf{MV-GLM...}}\\

\subsubsection{Update \(\mathbf{Q}\)}

\textcolor{red}{\textbf{MV-GLM...}}\\

\textcolor{red}{\textbf{Old results}}\\

\textcolor{red}{\textbf{Currently, the latent \(x_t\) estimation is not accurate enough. This may come from bad simulation/ slow convergence or other sources. See details in section \ref{problems}}}\\

There are two set of simulation examples. The first example is generated from LDS model directly, while in the second example the latents are generated directly without explicit specifying linear dynamics. In all of the following results, the covariance of process noise \(\mathbf{Q}\) is assumed to be block diagonal. For reference, see the code for \href{https://github.com/weigcdsb/state-space-clustering/blob/main/LDS/old_example/sim_LDS_noConstraint.m}{non-constrained} and \href{https://github.com/weigcdsb/state-space-clustering/blob/main/LDS/old_example/sim_LDS_diag.m}{diagonal} \(\mathbf{Q}\). But these two legacy versions assuming the loading is independent on cluster assignments. If we need to use any of one of them, make sure to replace the updates for loading to be cluster-related. The fitting results for all these three are similar, because the underlying true \(\mathbf{Q}\) is diagonal.

Before doing the clustering, I first assume the cluster labels are known to see the fitting performance. All the fitting results with labels are shown in means from iteration 50 to 100 (\textcolor{red}{\textbf{The iteration number is not enough to achieve stationary distribution}}).

\subsection{Simulation 1: Generate from LDS Directly}
In this simulation, there are 3 clusters with 10 neurons in each cluster. The dimension of latents in each cluster is 2. In the linear dynamics, the bias term \(\mathbf{b}\) is zero. There's no within-population interaction but has some weak between-population interactions. In other words, the linear dynamics matrix \(\mathbf{A}\) is roughly diagonal. The details of simulation can be found in the first section of the \href{https://github.com/weigcdsb/state-space-clustering/blob/main/LDS/lds_sample.m}{simulation 1}.

\subsubsection{Labeled Data: No Clustering}
The code for fitting is \href{https://github.com/weigcdsb/state-space-clustering/blob/main/LDS/lds_sample.m}{here}, and some results are in Figure \ref{fig:LDS labeled}.

\begin{figure}[h!]
	\makebox[\linewidth][c]{%
		\centering
		\subfigure[overall mean firing rate]{\label{fig:a}\includegraphics[width=0.55\textwidth]{image001.png}}%
		\subfigure[latents]{\label{fig:b}\includegraphics[width=0.55\textwidth]{image002.png}}%
		\subfigure[dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{image003.png}}%
	}
	\caption{LDS sample with labels}
	\label{fig:LDS labeled}
\end{figure}

The fitting for overall mean firing rate is perfect and the dynamics pattern is mostly recovered. The fitting of latents is OK (\textcolor{red}{\textbf{may not be accurate enough}}\\) up to scale. In other words, the model captures the oscillating pattern of latents.

But the latent patterns of these three clusters seems quite similar. That means the differences in the mean firing rates may be purely explained by the loading, i.e. \(\mathbf{d}\) and \(\mathbf{C}\). To see that, I set the number of cluster to be 1 and the results are shown in Figure \ref{fig:LDS labeled one cluster}.

\begin{figure}[h!]
	\makebox[\linewidth][c]{%
		\centering
		\subfigure[overall mean firing rate]{\label{fig:a}\includegraphics[width=0.7\textwidth]{image004.png}}%
		\subfigure[latents]{\label{fig:b}\includegraphics[width=0.7\textwidth]{image005.png}}%
	}
	\caption{LDS sample with labels, one cluster}
	\label{fig:LDS labeled one cluster}
\end{figure}

Again, the fitting for overall mean firing rate is perfect. This is why it's important to make the loading, \(\mathbf{d}\) and \(\mathbf{C}\), also be cluster-dependent (loading within clusters are correlated). If the loading only depends on neuron index and will not change for different clustering assignments, it's impossible to do clustering (at least in this case), since the loading is enough to capture all the patterns.

\subsubsection{Unlabeled Data: Clustering}
To give the full path of clustering, I show results in GIFs. Click to see the code for \href{https://github.com/weigcdsb/state-space-clustering/blob/main/LDS/lds_sample_MM.m}{FMM} and \href{https://github.com/weigcdsb/state-space-clustering/blob/main/LDS/lds_sample_DP.m}{DPMM}.

There are three results:
\begin{enumerate}
	\def\labelenumi{(\arabic{enumi})}
	\item
	Fit by FMM with \(J = 3\), starting from random cluster assignments:
	\href{https://github.com/weigcdsb/state-space-clustering/blob/main/results/gif/lds_samp_10_MM_above.gif}{GIF result 1}
	\item
	Fit by FMM with \(J = 3\), starting from all neurons in single cluster: 
	\href{https://github.com/weigcdsb/state-space-clustering/blob/main/results/gif/lds_samp_10_MM_below.gif}{GIF result 2}
	\item
	Fit by DPMM (\(\alpha = 10\)), starting from each neuron forms its own cluster:
	\href{https://github.com/weigcdsb/state-space-clustering/blob/main/results/gif/lds_samp_10_DP_above_10.gif}{GIF result 3}

\end{enumerate}
I didn't show the DPMM starting from single cluster. Since in my current implementation, the cluster generation is not efficient. In other words, the newly generated cluster will usually not be sampled and the algorithm usually gets stuck in 1 or 2 clusters.

This is a big problem... That means if we occasionally combine two clusters into one, we may never correct the mistake. Further, when the data grows, the number of cluster will be hard to grow. Fix that later.

\subsection{Simulation 2: Generate Latents, without Specifying Linear Dynamics}
In this simulation, there are again 3 clusters with 2 latents in each. Besides set 10 neurons in each cluster, I further simulate 50 neurons in each cluster to see performance in a larger scale. Now, the latents are generated directly without specifying the underlying linear dynamics of latents. However, each latent is generated independently, so the linear dynamics matrix \(\mathbf{A}\) should be roughly diagonal. The details of simulation can be found in the \href{https://github.com/weigcdsb/state-space-clustering/blob/main/LDS/unspecifiedA_sample.m}{code}.

\subsubsection{Labeled Data: No Clustering}
As in simulation 1, the data is fitted by the true cluster assignment (3 clusters) and forcing all neurons belong to single cluster. The code can be found \href{https://github.com/weigcdsb/state-space-clustering/blob/main/LDS/unspecifiedA_sample.m}{here}.
\begin{enumerate}
	\def\labelenumi{(\arabic{enumi})}
	\item
	Smaller dataset (10 neurons each):\\
	The results with true cluster assignment is shown in Figure \ref{fig:no A labeled 10}
	\begin{figure}[h!]
		\makebox[\linewidth][c]{%
			\centering
			\subfigure[overall mean firing rate]{\label{fig:a}\includegraphics[width=0.55\textwidth]{image006.png}}%
			\subfigure[latents]{\label{fig:b}\includegraphics[width=0.55\textwidth]{image007.png}}%
			\subfigure[dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{image008.png}}%
		}
		\caption{unspecified dynamics with labels, 10 neurons each}
		\label{fig:no A labeled 10}
	\end{figure}
	And results by forcing all neurons to be in single cluster (Figure \ref{fig:no A labeled one cluster 10})
	\begin{figure}[h!]
		\makebox[\linewidth][c]{%
			\centering
			\subfigure[overall mean firing rate]{\label{fig:a}\includegraphics[width=0.7\textwidth]{image009.png}}%
			\subfigure[latents]{\label{fig:b}\includegraphics[width=0.7\textwidth]{image010.png}}%
		}
		\caption{unspecified dynamics with labels, one cluster, 10 neurons each}
		\label{fig:no A labeled one cluster 10}
	\end{figure}

	Again, the fittings for overall mean firing rate are perfect in both cases and the dynamics is roughly diagonal. The oscillating pattern of fitted latents with true cluster index is not significant, because of plot scaling. The pattern is easier to see in the second fitting.

	\item
	Larger dataset (50 neurons each):\\
	The results with true cluster assignment is shown in Figure \ref{fig:no A labeled 50}
	\begin{figure}[h!]
		\makebox[\linewidth][c]{%
			\centering
			\subfigure[overall mean firing rate]{\label{fig:a}\includegraphics[width=0.55\textwidth]{image011.png}}%
			\subfigure[latents]{\label{fig:b}\includegraphics[width=0.55\textwidth]{image012.png}}%
			\subfigure[dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{image013.png}}%
		}
		\caption{unspecified dynamics with labels, 50 neurons each}
		\label{fig:no A labeled 50}
	\end{figure}
	And results by forcing all neurons to be in single cluster (Figure \ref{fig:no A labeled one cluster 50})
	\begin{figure}[h!]
		\makebox[\linewidth][c]{%
			\centering
			\subfigure[overall mean firing rate]{\label{fig:a}\includegraphics[width=0.7\textwidth]{image014.png}}%
			\subfigure[latents]{\label{fig:b}\includegraphics[width=0.7\textwidth]{image015.png}}%
		}
		\caption{unspecified dynamics with labels, one cluster, 50 neurons each}
		\label{fig:no A labeled one cluster 50}
	\end{figure}

	Now we can see the pattern more clearly.
\end{enumerate}

\subsubsection{Unlabeled Data: Clustering}
As previous, in each example, I show results from (1) FMM starting from random cluster assignment \(\left(J = 3\right) \), (2) FMM starting from single cluster and (3) DPMM starting from multiple clusters. The DPMM starting from single cluster need to be fixed later. These are code for
\href{https://github.com/weigcdsb/state-space-clustering/blob/main/LDS/unspecifiedA_sample_MM.m}{FMM} and \href{https://github.com/weigcdsb/state-space-clustering/blob/main/LDS/unspecifiedA_sample_DP.m}{DPMM}.

\begin{enumerate}
	\def\labelenumi{(\arabic{enumi})}
	\item
	Smaller dataset (10 neurons each):\\
	\begin{enumerate}
		\def\labelenumi{\alph{enumi}.}
		\item
		Fit by FMM with \(J = 3\), start from random cluster assignments:
		\href{https://github.com/weigcdsb/state-space-clustering/blob/main/results/gif/noA_10_MM_above.gif}{GIF result 4}
		\item
		Fit by FMM with \(J = 3\), start from single cluster:
		\href{https://github.com/weigcdsb/state-space-clustering/blob/main/results/gif/noA_10_MM_below.gif}{GIF result 5}
		\item
		Fit by DP (\(\alpha = 5\)), start from assuming each neuron forms its own cluster:
		\href{https://github.com/weigcdsb/state-space-clustering/blob/main/results/gif/noA_10_DP_above_5.gif}{GIF result 6}
	\end{enumerate}

	\item
	Larger dataset (50 neurons each):\\
	\begin{enumerate}
		\def\labelenumi{\alph{enumi}.}
		\item
		Fit by FMM with \(J = 3\), start from random cluster assignments:
		\href{https://github.com/weigcdsb/state-space-clustering/blob/main/results/gif/noA_50_MM_above.gif}{GIF result 7}
		\item
		Fit by FMM with \(J = 3\), start from single cluster:
		\href{https://github.com/weigcdsb/state-space-clustering/blob/main/results/gif/noA_50_MM_below.gif}{GIF result 8}
		\item
		Fit by DP (\(\alpha = 1\)), start from random assignment for 10 clusters:
		\href{https://github.com/weigcdsb/state-space-clustering/blob/main/results/gif/noA_50_DP_above1.gif}{GIF result 9}
	\end{enumerate}
\end{enumerate}

The performance is not bad. The neuron at top of each cluster cannot be clustered correctly by its nature: the signals are too weak to be clearly clustered.

\section{PROBLEMS} \label{problems}
This section shows what cause the inaccuracy in the latent estimation.

\subsection{Estimation of \(\mathbf{d}\) and \(\mathbf{C}\) alone}
At first, as guessed by Ian, this may be caused by bad estimation of loading (\(\mathbf{d}\) and \(\mathbf{C}\)). So I first turn off all others except for loading (related) parameters and set them to be true values. There are 2 versions: (1) update priors as in subsection \ref{loading prior} and (2) no update of priors, and just set the prior for each as the standard normal.
For the estimation with prior update, the results are mean from iteration 50 to 100 (Figure \ref{loading alone}).
\begin{figure}[h!]
	\makebox[\linewidth][c]{%
		\centering
		\subfigure[]{\includegraphics[width=0.6\textwidth]{image016.png}}%
		\subfigure[]{\includegraphics[width=0.6\textwidth]{image017.png}}%
	}
	\caption{estimation of loading}
	\label{loading alone}
\end{figure}

@@

I've checked update of (1) dynamics alone, (2) \(\mathbf{Q}\) and (3) \(\mathbf{Q}\) and \(\mathbf{x}_{t}\). All of them are fine. \textbf{So the problem must come from estimations of dynamics and \(\mathbf{x}_t\)}.

\subsection{Estimation of \(\mathbf{b}\), \(\mathbf{A}\) and \(\mathbf{x}_t\)}
I first estimate dynamics by assuming block diagonal \(\mathbf{Q}\) (as in subsection \ref{dynamics update}). When turning on (1) latents \(\mathbf{x}_t\) and (2) dynamics (\(\mathbf{b}\) and \(\mathbf{A}\)), the trace of \(\mathbf{b}\) and \(\mathbf{A}\) are (Figure \ref{trace of dynamics, blk}):\\
\begin{figure}[h!]
	\centering
	\includegraphics[width = .7\textwidth]{image022.png}
	\caption{trace of dynamics, block diagonal}
	\label{trace of dynamics, blk}
\end{figure}

The mixing is not good, and we may haven't achieve stationary distribution yet. Anyway, I average results from iteration 5000 to 10000, and the results are shown in Figure \ref{lat and dyn, blk}).
\begin{figure}[h!]
	\makebox[\linewidth][c]{%
		\centering
		\subfigure[latent]{\includegraphics[width=0.6\textwidth]{image023.png}}%
		\subfigure[dynamics]{\includegraphics[width=0.6\textwidth]{image024.png}}%
	}
	\caption{estimation of latents and dynamics, block diagonal}
	\label{lat and dyn, blk}
\end{figure}

>>>>

To facilitate derivation in Gibbs sampler, write the linear dynamics of
\(\mathbf{x}_{t}\) in MV-GLM form, i.e.

\[\mathbf{x}_{t + 1}^{'} = \begin{pmatrix}
	1 & \mathbf{x}_{t}^{'} \\
\end{pmatrix}\begin{pmatrix}
	\mathbf{b}^{'} \\
	\mathbf{A}^{'} \\
\end{pmatrix} + \mathbf{\epsilon}_{i}^{'}\]

, where \(\mathbf{\epsilon} \sim N(0,\mathbf{Q})\). Then if we stack the
latent by row, the problem is reduced to a multivariate general linear model (MV-GLM) problem:

\[\begin{pmatrix}
	\mathbf{x}_{2}^{'} \\
	\mathbf{x}_{3}^{'} \\
	\vdots \\
	\mathbf{x}_{T}^{'} \\
\end{pmatrix} = \begin{pmatrix}
	1 & \mathbf{x}_{1}^{'} \\
	1 & \mathbf{x}_{2}^{'} \\
	\vdots & \vdots \\
	1 & \mathbf{x}_{T - 1}^{'} \\
\end{pmatrix}\begin{pmatrix}
	\mathbf{b}^{'} \\
	\mathbf{A}^{'} \\
\end{pmatrix} + \mathbf{E}\]

, where
\(\mathbf{E} = \left( \mathbf{\epsilon}_{1},\ldots,\mathbf{\epsilon}_{T - 1} \right)^{'}\).

@@

The conditional priors for these parameters:
\begin{enumerate}
	\def\labelenumi{(\arabic{enumi})}
	\item
	Latent vectors \(\left\{ \mathbf{x}_{t} \right\}_{t=1}^T\): the conditional prior is
	defined by
	\[\mathbf{x}_{1} \sim N(\mathbf{x}_{0},\mathbf{Q}_{0})\]
	\[\mathbf{x}_{t + 1}|\mathbf{x}_{t} \sim N(\mathbf{A}\mathbf{x}_{t} + \mathbf{b},\mathbf{Q})\]
	\item
	Initials \(\mathbf{x}_{0}\): assume there are \(J\) clusters,
	\[\mathbf{x}_{0} \sim N(\bm{\mu}_{\mathbf{x}_{00}},\ \mathbf{\Sigma}_{\mathbf{x}_{00}})\]
	, where \(\bm{\mu}_{\mathbf{x}_{00}} = \mathbf{0}_{\text{Jp}}\) and \(\mathbf{\Sigma}_{\mathbf{x}_{00}} = \mathbf{I}_{\text{Jp}}\)
	\item
	Linear mapping (loading) for latent vectors
	\(\left\{ d_{i} \right\}_{i = 1}^{N}\) and
	\(\left\{ \mathbf{c}_{i} \right\}_{i = 1}^{N}\):
	\[\left( d_{i},\mathbf{c}'_{i} \right)' \sim N(\bm{\mu}_{\text{dc}}^{\left( z_{i} \right)},\mathbf{\Sigma}_{\text{dc}}^{(z_{i})})\]
	\item
	Mean and covariance for loading in each cluster
	Mean and covariance for loading in each cluster
	\(\left\{ \bm{\mu}_{\text{dc}}^{(j)} \right\}_{j}\) and
	\(\left\{ \mathbf{\Sigma}_{\text{dc}}^{(j)} \right\}_{j}\):
	\[\bm{\mu}_{\text{dc}}^{(j)} \sim N(\mathbf{\delta}_{dc0},\mathbf{T}_{dc0})\]
	, where \(\bm{\delta}_{dc0} = \mathbf{0}_{p + 1}\) and
	\(\mathbf{T}_{dc0} = \mathbf{I}_{p + 1}\)

	\[\mathbf{\Sigma}_{\text{dc}}^{(j)} \sim W^{- 1}\left( \Psi_{dc0},\nu_{dc0} \right)\]
	, where \(\nu_{dc0} = p + 1 + 2\) and
	\(\Psi_{dc0} = \mathbf{I}_{p + 1} \times 10^{-4}\)

	\item
	Linear dynamics for latent vectors: \(\mathbf{A}\), \(\mathbf{b}\) and process noise \(\mathbf{Q}\).\\ 
	Denote \(\mathbf{F} = \begin{pmatrix}
		\mathbf{b}^{'} \\
		\mathbf{A}^{'} \\
	\end{pmatrix}\), \(\mathbf{f} = vec(\mathbf{F})\)

	\begin{align*}
		P\left( \mathbf{f},\ \mathbf{Q} \right) &= P(\mathbf{Q})P(\mathbf{f}|\mathbf{Q})\\
		\mathbf{Q} &\sim W^{- 1}(\Psi_{\mathbf{Q}_{0}},\nu_{\mathbf{Q}_{0}})\\
		\mathbf{f} &\sim N(\mathbf{f}_0, \mathbf{Q}\bigotimes \bm{\Lambda}_0^{-1})
	\end{align*}

	, where \(\mathbf{f}_0 = vec(\mathbf{F}_0)\).If the number of cluster is \(J\), \(\nu_{\mathbf{Q}_{0}} = Jp + 2\),
	\(\Psi_{\mathbf{Q}_{0}} = \mathbf{I}_{\text{Jp}} \times 10^{- 4}\). (To make the mean of \(\mathbf{Q}\) loosely centered around
	\(\mathbf{I}_{Jp} \times 10^{- 4}\)), \(\mathbf{F}_0 = \begin{pmatrix}
		\mathbf{0}'_{Jp} \\
		\mathbf{I}_{\text{Jp}} \\
	\end{pmatrix}\) and \(\bm{\Lambda}_0 = \mathbf{I}_{Jp + 1}\)

@@

\subsubsection{Update \(\left\{ d_{i} \right\}_{i = 1}^{N}\) and
	\(\left\{ \mathbf{c}_{i} \right\}_{i = 1}^{N}\)}
To update efficiently, use Laplace approximation and the error is independent for each row (i.e. update things row by row). Denote
\(\left( d_{i},\mathbf{c}'_{i} \right)' = \bm{\zeta}_{i} \in \mathbb{R}^{p + 1}\)
and \(\left( 1,{\mathbf{x}_{t}'^{\left( z_{i} \right)}} \right) = {{\widetilde{\mathbf{x}}}_{t}'^{\left( z_{i} \right)}}\).
\begin{align*}
	P\left( \bm{\zeta}_{i} \middle| \mathbf{y}_{i},\ \left\{ \mathbf{x}_{t}^{\left( z_{i} \right)} \right\}_{t = 1}^{T},\ldots \right) &= \exp{f\left( \bm{\zeta}_{i} \right)} \approx N\left( \bm{\zeta}_{i}|\bm{\mu}_{\bm{\zeta}_{i}},\mathbf{\Sigma}_{\bm{\zeta}_{i}}\  \right)\\
	\frac{\partial f}{\partial\bm{\zeta}_{i}} &= \frac{\partial l}{\partial\bm{\zeta}_{i}} - {\mathbf{\Sigma}_{\text{dc}}^{\left( z_{i} \right)}}^{- 1}\left( \bm{\zeta}_{i} - \bm{\mu}_{\text{dc}}^{(z_{i})} \right) = \left\lbrack \sum_{t = 1}^{T}{\widetilde{\mathbf{x}}}_{t}^{\left( z_{i} \right)}\left( y_{\text{it}} - \lambda_{\text{it}} \right) \right\rbrack - {\mathbf{\Sigma}_{\text{dc}}^{\left( z_{i} \right)}}^{- 1}\left( \bm{\zeta}_{i} - \bm{\mu}_{\text{dc}}^{(z_{i})} \right)\\
	\frac{\partial^{2}f}{\partial\bm{\zeta}_{i}\partial\bm{\zeta}'_{i}} &= \frac{\partial^{2}l}{\partial\bm{\zeta}_{i}\partial\bm{\zeta}'_{i}} - {\mathbf{\Sigma}_{\text{dc}}^{\left( z_{i} \right)}}^{- 1} = -\left\lbrack \sum_{t = 1}^{T}{\lambda_{\text{it}}{\widetilde{\mathbf{x}}}_{t}^{\left( z_{i} \right)}{{\widetilde{\mathbf{x}}}_{t}'^{\left( z_{i} \right)}}} \right\rbrack - {\mathbf{\Sigma}_{\text{dc}}^{\left( z_{i} \right)}}^{- 1}
\end{align*}
, where \(l\) is Poisson log-likelihood. Then use Newton-Raphson to find
\(\bm{\mu}_{\bm{\zeta}_{i}} = \text{argmax}_{\bm{\zeta}_{i}}\left( f\left( \bm{\zeta}_{i} \right) \right)\)
and \(\mathbf{\Sigma}_{\bm{\zeta}_{i}} = -\left\lbrack \frac{\partial^{2}f}{\partial\bm{\zeta}_{i}\partial\bm{\zeta}'_{i}}\left.  \right|_{\bm{\zeta}_{i} = \bm{\mu}_{\bm{\zeta}_{i}}} \right\rbrack^{-1}\). If initial as the previous step fits fails, simply use prior mean of \(\bm{\mu}_{\bm{\zeta}_{i}}\), i.e. \(\bm{\delta}_{dc0}\).

DETOUR: we can write things in the form of MV-(Poisson)-GLM as follows (in cluster j with \(n_j\) observations/ neurons, with corresponding loading be \(\mathbf{d}_{j}\) and \(\mathbf{C}_{j}\)):

\[\log\begin{pmatrix}
	\lambda_{11} & \cdots & \lambda_{1n_{j}} \\
	\vdots & \ddots & \vdots \\
	\lambda_{T1} & \cdots & \lambda_{Tn_{j}} \\
\end{pmatrix} = \begin{pmatrix}
	1 & \mathbf{x}_{1}^{'} \\
	\vdots & \vdots \\
	1 & \mathbf{x}_{T}^{'} \\
\end{pmatrix}\begin{pmatrix}
	\mathbf{d}_{j}^{'} \\
	\mathbf{C}_{j}^{'} \\
\end{pmatrix} + \mathbf{E}_j\]

We can further organize things column-wise. let \(\begin{pmatrix}
	1 & \mathbf{x}_{1}^{'} \\
	\vdots & \vdots \\
	1 & \mathbf{x}_{T}^{'} \\
\end{pmatrix} = \mathbf{G}\), then the design matrix and corresponding log-mean response are \(\begin{pmatrix}
\mathbf{G} & \mathbf{0} & \cdots & \mathbf{0} \\
\mathbf{0} & \mathbf{G} & \cdots & \mathbf{0} \\
\vdots & \vdots & \ddots & \vdots \\
\mathbf{0} & \cdots & \cdots & \mathbf{G} \\
\end{pmatrix}\) and \({\log\begin{pmatrix}
	\lambda_{11}^{(j)} & \cdots & \lambda_{T1}^{(j)} & \lambda_{12}^{(j)} & \cdots & \lambda_{T2}^{(j)} & \cdots & \lambda_{Tn_{j}}^{(j)} \\
\end{pmatrix}}'\).

However, when I use the Laplace approximation and NR, because of the block-diagonal structure of design matrix, the error is always independent observation by observation. This is equivalent to observation-wise update as above.

\textbf{Maybe I did something wrong?}
The point to say this is to make loading within each cluster depends on each other. Check these two later(\href{https://www.tandfonline.com/doi/full/10.1080/03610926.2012.743565?journalCode=lsta20}{ref1} and \href{https://www.tandfonline.com/doi/full/10.1080/02664763.2021.1877637?src=recsys}{ref2}).

Now I do it in another way: make the prior cluster-dependent...

@@

Purpose: to make loading within each cluster depends on each other, and this will help with clustering. As above, denote \(\left( d_{i},\mathbf{c}'_{i} \right)' = \bm{\zeta}_{i} \in \mathbb{R}^{p + 1}\).
\begin{enumerate}
	\def\labelenumi{(\arabic{enumi})}
	\item
	Mean \(\left\{ \bm{\mu}_{\text{dc}}^{(j)} \right\}_{j}\): by
	conjugacy, \(\bm{\mu}_{\text{dc}}^{(j)} \sim N\left( \bm{\delta}_{\text{dc}},\mathbf{T}_{\text{dc}}\  \right)\)
	\begin{align*}
		\mathbf{T}_{\text{dc}}^{- 1} &= \left( \mathbf{T}_{dc0}^{- 1} + n_{j}{\mathbf{\Sigma}_{\text{dc}}^{(j)}}^{- 1} \right)^{- 1}\\
		\bm{\delta}_{\text{dc}} &= \mathbf{T}_{\text{dc}}\left( \mathbf{T}_{dc0}^{- 1}\bm{\delta}_{dc0} + {\mathbf{\Sigma}_{\text{dc}}^{(j)}}^{- 1}\sum_{i:z_{i} = j}^{}\bm{\zeta}_{i} \right)
	\end{align*}
	\item
	Covariance \(\left\{ \mathbf{\Sigma}_{\text{dc}}^{(j)} \right\}_{j}\): by conjugacy, \(\mathbf{\Sigma}_{\text{dc}}^{(j)} \sim W^{- 1}\left( \Psi_{\text{dc}},\nu_{\text{dc}} \right)\)
	\begin{align*}
		\nu_{\text{dc}} &= n_{j} + \nu_{dc0}\\
		\Psi_{\text{dc}} &= \Psi_{dc0} + \sum_{i:z_{i} = j}^{}{\left( \bm{\zeta}_{i} - \bm{\mu}_{\text{dc}}^{(j)} \right)\left( \bm{\zeta}_{i} - \bm{\mu}_{\text{dc}}^{(j)} \right)'}
	\end{align*}
\end{enumerate}

\subsubsection{Update \(\mathbf{A}\), \(\mathbf{b}\) and \(\mathbf{Q}\)} 
Here, I only give the update for full \(\mathbf{Q}\). If we want to assume block-diagonal or diagonal structure of \(\mathbf{Q}\), just update things cluster-by-cluster or latent-by-latent.

As shown before, the problem is the usual Bayesian MV-GLM problem. Denote \(\mathbf{F} = \begin{pmatrix}
	\mathbf{b}^{'} \\
	\mathbf{A}^{'} \\
\end{pmatrix}\), \(\mathbf{f} = vec(\mathbf{F})\), \(\mathbf{Y}_\mathbf{bA} = (\mathbf{x}_{2}, \mathbf{x}_{3} \ldots, \mathbf{x}_{T})'\) and \(\mathbf{X}_\mathbf{bA} = \begin{pmatrix}
1 & \mathbf{x}_{1}^{'} \\
1 & \mathbf{x}_{2}^{'} \\
\vdots & \vdots \\
1 & \mathbf{x}_{T - 1}^{'} \\
\end{pmatrix}\)

\[\mathbf{Y}_\mathbf{bA} = \mathbf{X}_\mathbf{bA}\mathbf{F}+ \mathbf{E}\]

, where
\(\mathbf{E} = \left( \mathbf{\epsilon}_{1},\ldots,\mathbf{\epsilon}_{T - 1} \right)^{'}\) and \(\mathbf{\epsilon} \sim N(0,\mathbf{Q})\). Then posteriors are
\begin{align*}
	\mathbf{Q}|\mathbf{Y}_{\mathbf{\text{bA}}},\mathbf{X}_{\mathbf{\text{bA}}},\ldots &\sim W^{-1}(\Psi_{\mathbf{Q}}, \nu_{\mathbf{Q}})\\
	\mathbf{f}|\mathbf{Y}_{\mathbf{\text{bA}}},\mathbf{X}_{\mathbf{\text{bA}}},\mathbf{Q},\ldots &\sim N(vec(\mathbf{F}_\mathbf{bA}),  \mathbf{Q}\bigotimes \bm{\Lambda}_\mathbf{bA}^{-1})
\end{align*}
, with parameters be:
\begin{align*}
	\Psi_{\mathbf{Q}} &= \Psi_{\mathbf{Q}_{0}} + (\mathbf{Y}_\mathbf{bA} - \mathbf{X}_\mathbf{bA}\mathbf{F}_\mathbf{bA})'(\mathbf{Y}_\mathbf{bA} - \mathbf{X}_\mathbf{bA}\mathbf{F}_\mathbf{bA}) + (\mathbf{F}_\mathbf{bA} - \mathbf{F}_0)'\bm{\Lambda}_0(\mathbf{F}_\mathbf{bA} - \mathbf{F}_0)\\
	\nu_{\mathbf{Q}} &= \nu_{\mathbf{Q}_{0}} + T - 1\\
	\mathbf{F}_\mathbf{bA} &= (\mathbf{X}'_\mathbf{bA}\mathbf{X}_\mathbf{bA} + \bm{\Lambda}_0)^{-1}(\mathbf{X}'_\mathbf{bA}\mathbf{Y}_\mathbf{bA} + \bm{\Lambda}_0\mathbf{F}_0)\\
	\bm{\Lambda}_\mathbf{bA} &= \mathbf{X}'_\mathbf{bA}\mathbf{X}_\mathbf{bA} + \bm{\Lambda}_0
\end{align*}

\section{Model Check [Problems for \(\mathbf{b}\) and \(\mathbf{A}\) partially solved]} \label{problems}

\subsection{Estimation of \(\mathbf{d}\) and \(\mathbf{C}\) alone}
First, turn off all others except for loading (related) parameters and set them to be true values. There are 2 versions: (1) update priors as in subsection \ref{loading prior} and (2) no update of priors, and just set the prior for each as the standard normal.
For the estimation with prior update, the results are mean from iteration 50 to 100 (Figure \ref{loading alone}).
\begin{figure}[h!]
	\makebox[\linewidth][c]{%
		\centering
		\subfigure[]{\includegraphics[width=0.6\textwidth]{image016.png}}%
		\subfigure[]{\includegraphics[width=0.6\textwidth]{image017.png}}%
	}
	\caption{estimation of loading}
	\label{loading alone}
\end{figure}

@@

\subsection{Estimation of \(\mathbf{b}\), \(\mathbf{A}\), \(\mathbf{Q}\) and \(\mathbf{x}_t\)}
I show the results of block-diagonal (Figure \ref{bAQX_blkDiag}) and full (Figure \ref{bAQX_full}) \(\mathbf{Q}\). These results are from 2000 MCMC samples. The latent and dynamics are from average from 500 to 2000 iterations.

\begin{figure}[h!]
	\makebox[\linewidth][c]{%
		\centering
		\subfigure[trace]{\label{fig:a}\includegraphics[width=0.55\textwidth]{bAX_blkDiag_trace.png}}%
		\subfigure[latent]{\label{fig:b}\includegraphics[width=0.55\textwidth]{bAX_blkDiag_latent.png}}%
		\subfigure[dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{bAX_blkDiag_A.png}}%
	}
	\caption{sampling of $\mathbf{b}$, $\mathbf{A}$, $\mathbf{Q}$ and $\mathbf{x}_t$, block-diagonal process noise}
	\label{bAQX_blkDiag}
\end{figure}

\begin{figure}[h!]
	\makebox[\linewidth][c]{%
		\centering
		\subfigure[trace]{\label{fig:a}\includegraphics[width=0.55\textwidth]{bAX_full_trace.png}}%
		\subfigure[latent]{\label{fig:b}\includegraphics[width=0.55\textwidth]{bAX_full_latent.png}}%
		\subfigure[dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{bAX_full_A.png}}%
	}
	\caption{sampling of $\mathbf{b}$, $\mathbf{A}$, $\mathbf{Q}$ and $\mathbf{x}_t$, full process noise}
	\label{bAQX_full}
\end{figure}

Both block-diagonal and full looks fine, but the mixing of  full process noise version looks better.
